---
title: "Practical Machine Learning Project - Report"
author: "Harkishan Grewal"
date: "12/31/2020"
output:
  html_document
---

### Load Required Libraries

```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(knitr)
```


# Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible 
to collect a large amount of data about personal activity relatively 
inexpensively. These type of devices are part of the quantified self movement 
– a group of enthusiasts who take measurements about themselves regularly to 
improve their health, to find patterns in their behavior, or because they are 
tech geeks. One thing that people regularly do is quantify how much of a 
particular activity they do, but they rarely quantify how well they do it.

In this project, I will use the data from accelerometers on the belt, 
forearm, arm, and dumbell of 6 participants. The participants were asked to 
perform barbell lifts correctly and incorrectly in 5 different ways.
More information is available from the website here:
http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har
(see the section on the Weight Lifting Exercise Dataset).


# Data

The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har.
Please cite the source if their data is used.


## Data Preprocessing

### Download the Data

Here I download the data, if it is not downloaded already.

```{r}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "./pml-training.csv"
testFile  <- "./pml-testing.csv"

# Download training data file if not downloaded already
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile=trainFile, method="curl")
}

# Download testing data file if not downloaded already
if (!file.exists(testFile)) {
  download.file(testUrl, destfile=testFile, method="curl")
}
```


### Read the Data

After downloading the data from the data source, I read the data from the two 
files.

```{r}
trainRaw <- read.csv("./pml-training.csv", na.strings = c("NA","#DIV/0!",""))
testRaw <- read.csv("./pml-testing.csv", na.strings=c("NA","#DIV/0!",""))
```

As shown below, the training dataset has 19622 observations and 160 variables.
The testing dataset has 20 observations and 160 variables.

```{r}
dim(trainRaw)
dim(testRaw)
```


### Split the Data

Next, I prepare the data for prediction by splitting the training data into a 
training set and a testing set. The ratio of training to testing data is 70-30. 

```{r}
set.seed(2021)
train <- createDataPartition(trainRaw$classe, p = 0.7, list = FALSE)
trainData <- trainRaw[train, ]
testData <- trainRaw[-train, ]
```

The dimensions of the datasets after the split are shown below.

```{r}
dim(trainData)
dim(testData)
```

### Clean the Data

I remove near-zero-variance (NZV) variables from train and test data.

```{r}
nzv <- nearZeroVar(trainData, saveMetrics = TRUE)
trainData <- trainData[, nzv$nzv == FALSE]

nzv <- nearZeroVar(testData, saveMetrics = TRUE)
testData <- testData[, nzv$nzv == FALSE]
```

The dimensions of the datasets after near-zero-variance variables are shown 
below.

```{r}
dim(trainData)
dim(testData)
```

I clean the data by removing variables that contain missing values.

```{r}
trainData <- trainData[c(-1)]
tempTrain <- trainData
for(i in 1:length(trainData)){
    if(sum(is.na(trainData[, i])) / nrow(trainData) >= 0.7){
        for(j in 1:length(tempTrain)){
            if(length(grep(names(trainData[i]), names(tempTrain)[j])) == 1){
                tempTrain <- tempTrain[, -j]
            }
        }
    }
}

trainData <- tempTrain
rm(tempTrain)
```

The dimensions of the datasets after removing variables containing missing 
values are shown below.

```{r}
dim(trainData)
dim(testData)
```

Transform the test data set and the raw test data set to have same dimensions 
as the training set after removing NA's.

```{r}
header <- colnames(trainData)
header1 <- colnames(trainData[, -58])
testData <- testData[header]
testRaw <- testRaw[header1]
```

The dimensions of the datasets after transform are shown below.

```{r}
dim(trainData)
dim(testData)
dim(testRaw)
```

Next, I coerce the data into the same data type.

```{r}
for (i in 1:length(testRaw)){
    for(j in 1:length(trainData)){
        if(length(grep(names(trainData[i]), names(testRaw)[j])) == 1){
            class(testRaw[j]) <- class(trainData[i])
        }
    }
}

testRaw <- rbind(trainData[2, -58] , testRaw)
testRaw <- testRaw[-1, ]
```


## Build a Model

I will use decision trees, random forests, and generalized boosted model 
algorithms to build a model.

### Decision Trees Model

Now, I create a decision tree model and plot the tree as a dendogram.

```{r}
dtm <- rpart(classe ~ ., data = trainData, method = "class")
fancyRpartPlot(dtm)
```

Next, I test the classification tree model on the test data.

```{r}
predictDTM <- predict(dtm, testData, type = "class")
cmDTM <- confusionMatrix(predictDTM, as.factor(testData$classe))
cmDTM

plot(cmDTM$table, col = cmDTM$byClass,
     main = paste("Decision Tree - Accuracy: ", round(cmDTM$overall['Accuracy'], 4)))
```


### Random Forest

Now, I create a random forest model.

```{r, cache=TRUE}
trainData$classe = factor(trainData$classe)
rf <- randomForest(classe ~ ., data = trainData)
predictRF <- predict(rf, testData, type = "class")
cmRF <- confusionMatrix(predictRF, as.factor(testData$classe))
cmRF

plot(cmRF$table, col = cmRF$byClass,
     main = paste("Random Forest - Accuracy: ", round(cmRF$overall['Accuracy'], 4)))
```


### Generalized Boosted Regression

Lastly, I create a generalized boosted regression model.

```{r, cache=TRUE}
control <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
gbm <- train(classe ~ ., data = trainData, method = "gbm", trControl = control,
                 verbose = FALSE)

predictGBM <- predict(gbm, testData)
cmGBM <- confusionMatrix(predictGBM, as.factor(testData$classe))
cmGBM

plot(cmGBM$table, col = cmGBM$byClass,
     main = paste("Generalized Boosted Regression - Accuracy: ", round(cmGBM$overall['Accuracy'], 4)))
```


## Test Data

Since Random Forest model had the highest accuracy, I will use it to predict 
the test data from the source. The model has an accuracy of **0.9995** so the out-of-sample rate is 100 - 99.95 = **0.05 %**. 

```{r}
predict <- predict(rf, testRaw, type = "class")
predict
```